---
title: "Homework 13"
format: html
---

Code for Homework 13, GLMs

```{r}
######HW 13
###Jackson D'Elia
###April 15

getwd()
###


#load libraries, load data. Using bee disease data wide.

library(tidyverse)
library(lubridate)
library(lme4)
library(car)

bee_dat <- read_csv("Burnham_field_data_pathogens_wide.csv")

bee_dat <- bee_dat %>%
  mutate(
    sampling_date = mdy(sampling_date),
    site_code = factor(site_code),
    field_id = factor(field_id),
    bee_caste = factor(bee_caste),
    bombus_spp = factor(bombus_spp),
    host_plant = factor(host_plant),
    sampling_event = factor(sampling_event),
    sampling_event_num = as.numeric(as.character(sampling_event)),
    log10_BQCV_load = log10(BQCV_pathogen_load + 1),
    log10_DWV_load = log10(DWV_pathogen_load + 1),
    log10_Nosema_load = log10(Nosema_pathogen_load + 1)
  )

glimpse(bee_dat)

df_filtered <- bee_dat[bee_dat$log10_Nosema_load > 0 & bee_dat$log10_BQCV_load > 0, ]

########
#MODEL 1
########
g_Nosema_by_field <- glm(
  log10_Nosema_load ~ field_id,
  data = df_filtered)

by_field <- Anova(g_Nosema_by_field)
print(by_field)
#Nosema load by field where sample was collected.
#this asks if location is a factor in nosema infection load.
#hypothesis: Nosema can be transmitted via flower visitation. Nosema load will differ across sites as some places will be disease 'hotspots'


#Graph of Nosema load by field site
qplot(
  x = field_id,
  y = log10_Nosema_load,
  data = df_filtered) +
geom_boxplot(method = "lm", se = TRUE)



```

The result is statistically signifigant with a p value of 0.031, suggesting that Nosema load is influenced by field id, meaning some sites result in specemins with higher loads than other sites.
there is a high amount of variation and bees are mobile between sites, so this conclusion might be presumptuous.

```{r}
########
#MODEL 2
########
g_bqcv_site <- lmer(
  log10_Nosema_load ~ bombus_spp + (1 | sampling_event),
  data = df_filtered)

Anova(g_bqcv_site)
```

A model of nosema load by bee species, taking into account random effects of sampling event.
This asks if different species are more susceptible to nosema infection


This model does not yeild a statistically signifigant result, suggesting that Nosema load is not influenced by bee species. 
# Nosema seems to infect bombus species equally

```{r}

#####################
#graphing model 2
#####################
df_filtered <- bee_dat[bee_dat$log10_Nosema_load > 0 & bee_dat$log10_BQCV_load > 0, ]


g2 <- ggplot(data=df_filtered) +
          aes(x=log10_Nosema_load) +
          geom_bar(fill="tomato",color="black")+ 
          theme(legend.position="none")
  theme(legend.position = "top")
  theme(legend.position = c(.2, .8))
print(g2)            
```

Model of nosema load, a non-gaussian distribution. 
# bees tend to have lower Nosema load, since the pathogen is harmful to the insect
Therefore we see a large amount of lower loads due to a survivorship bias

